\documentclass[conference]{IEEEtran}
\usepackage{gvv-book}
\usepackage{gvv}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{listings}
\usepackage{xcolor}
%\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{float}
\usepackage{booktabs}
\usepackage{url}
\usepackage{circuitikz}
\usepackage{hyperref}
\usepackage{tabularx}
\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue}\bfseries,
  commentstyle=\color{green!40!black},
  stringstyle=\color{red},
  xleftmargin=1.5em,
  frame=single,
  breaklines=true,
  breakatwhitespace=true,
  tabsize=2,
  showstringspaces=false,
  captionpos=b,
  keepspaces=true
}
\def\UrlBreaks{\do\/\do-\do\_\do\.}

\title{ESP-Flasher: Wireless Arduino and ESP32 Firmware Deployment from a Android Environment}

\author{
\IEEEauthorblockN{Darsh Gajare and Nishid Khandagre}
\IEEEauthorblockA{
Department of Electrical Engineering\\
IIT Hyderabad\\
Email: EE25BTECH11020@iith.ac.in, EE25BTECH11043@iith.ac.in}
}

\begin{document}

\maketitle

\begin{abstract}
Uploading firmware to microcontrollers from Android devices running Termux is
non-trivial because Android denies user-space processes direct USB access.
This paper presents ESP-Flasher, a two-tier wireless firmware deployment
system that eliminates the need for USB access during development.
In the first tier, an ESP32 acts as a transparent WiFi-to-UART bridge, relaying
the STK500v1 programming protocol from \texttt{avrdude} on the phone to the
Optiboot bootloader of an Arduino UNO over TCP port~328.
In the second tier, the same ESP32 doubles as a WiFi access-point and
HTTP firmware store, pushing verified binary images to a second ESP32 target
via raw TCP with MD5 integrity checking, bypassing the mDNS limitations of
Arduino OTA.
Together the two tiers form a fully wireless, USB-free firmware update chain
that is reproducible entirely from a command-line Termux environment.
\end{abstract}

\begin{IEEEkeywords}
ESP32, Arduino, Over-The-Air Update, WiFi-UART Bridge, STK500, PlatformIO,
Termux, Wireless Programming, Embedded Systems, LittleFS.
\end{IEEEkeywords}

%---------------------------------------------------------------------
\section{Introduction}
%---------------------------------------------------------------------

Wireless firmware deployment has attracted sustained interest across
embedded-systems research and practice.
The \textit{esp-link} project~\cite{esplink} demonstrated that an ESP8266
can host an integrated STK500v1 programmer together with a tiny HTTP server,
enabling AVR reprogramming over WiFi from any browser.
Lima and collaborators extended this concept to the three hardware UARTs of
the ESP32, providing a transparent TCP-to-serial bridge that forwards serial
data between a WiFi socket and the attached microcontroller~\cite{esp32bridge}.
On the OTA side, Espressif's official ESP-IDF provides a dual-partition update
scheme in which a new firmware image is written to the inactive OTA partition,
verified, and booted, guaranteeing that a failed update cannot brick the
device~\cite{espressifota}.
Joyce's \textit{esp32FOTA} library~\cite{esp32fota} further automates this
pipeline by polling a JSON manifest on a remote server and applying
semantic-version comparisons before initiating a download.
Mischianti~\cite{mischianti} surveys HTTP-based self-update patterns for
ESP32, highlighting the use of MD5 checksums as a lightweight integrity
mechanism and the role of LittleFS for intermediate firmware storage.
The ElegantOTA library~\cite{elegantota} packages the same HTTP update flow
into a ready-made web-server interface, while Flashgamer~\cite{flashgamer}
documents the low-level mechanics of chunked OTA streaming and partition
toggling in the ESP-IDF framework.

Despite this body of work, two practical gaps persist.
First, none of the WiFi-to-UART bridging solutions address the specific
constraint of a \emph{USB-restricted Android host}: prior art assumes a
desktop or laptop as the compile-and-upload machine, where \texttt{avrdude}
can open a COM port directly.
On Android with Termux, no such access is available, so the socket-based
upload path of \texttt{avrdude} (\texttt{-P socket://\textit{IP}:\textit{port}})
must carry the entire STK500 exchange over TCP---a configuration not discussed
in any of the references above.
Second, existing ESP-to-ESP OTA workflows either rely on mDNS service
discovery (which is unreliable on direct AP networks) or depend on external
cloud infrastructure.
Neither approach suits a fully air-gapped, access-point-local development
environment.

This paper makes two contributions that fill these gaps.
First, it characterises and resolves the bootloader timing problem introduced
by the DTR-less TCP connection: because \texttt{avrdude} cannot assert DTR
over a socket, the Optiboot bootloader's one-second window expires before
synchronisation is achieved.
We show that inserting a 100\,\textmu F capacitor between Arduino RST and GND
extends the reset window to 1--3\,s, making reliable STK500 programming over
TCP feasible.
Second, it presents a direct-IP HTTP OTA scheme in which one ESP32 acts as
both access point and firmware server, pushing a stored binary to a second
ESP32 over raw TCP with an \texttt{x-md5} header---requiring no cloud
infrastructure, no mDNS, and no laptop.
The complete system is operated entirely from a PlatformIO-based Termux
workflow on an Android phone.

%---------------------------------------------------------------------
\section{Uploading Code to Arduino Using ESP32 Wirelessly}
%---------------------------------------------------------------------

\subsection{Overview}

The phone compiles the sketch using PlatformIO and invokes \texttt{avrdude},
which opens a TCP socket to the ESP32 on port~328.
The ESP32 forwards all bytes transparently between the socket and UART2,
relaying the STK500 protocol to the Arduino bootloader.

\begin{table}[H]
\centering
\caption{System Components -- Chapter I}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{ll}
\toprule
\textbf{Component} & \textbf{Role} \\
\midrule
Phone (Termux + Debian) & Compile via PlatformIO; upload via \texttt{avrdude} \\
ESP32 NodeMCU & WiFiServer TCP:328; UART2 relay at 115200\,baud \\
Arduino UNO (ATmega328P) & Optiboot bootloader at 115200\,baud \\
\bottomrule
\end{tabular}}
\end{table}

\subsection{Hardware Connections}

\begin{table}[H]
\centering
\caption{ESP32--Arduino Pin Connections}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{llll}
\toprule
\textbf{ESP32 Pin} & \textbf{Arduino Pin} & \textbf{Dir.} & \textbf{Note} \\
\midrule
5V            & VIN         & $\rightarrow$     & Power via onboard regulator \\
GND           & GND         & $\leftrightarrow$ & Common ground --- essential \\
P17 (TX2)     & Pin 0 (RX)  & $\rightarrow$     & Direct wire \\
P16 (RX2)     & Pin 1 (TX)  & $\leftarrow$      & Via 3$\times$10\,k$\Omega$ divider \\
P4            & RST         & $\rightarrow$     & Reset control \\
\bottomrule
\end{tabular}}
\end{table}

\subsubsection{Pin Wiring}
The TX/RX lines are cross-connected between the ESP32 UART2 and the Arduino
hardware serial pins.

\subsubsection{Voltage Divider}
Arduino TX outputs 5\,V logic; ESP32 GPIO pins support 3.3\,V.
A resistor voltage divider on the P16 (RX2) line is required:
\begin{equation}
V_{\text{out}} = V_{\text{in}} \times \frac{R_2}{R_1+R_2}
              = 5\,\text{V} \times \frac{10\,\text{k}}{30\,\text{k}}
              = 3.3\,\text{V}
\end{equation}

\subsubsection{Reset Capacitor}
A 100\,\textmu F electrolytic capacitor connected between Arduino RST ($+$) and
GND ($-$) extends the bootloader window to 1--3\,seconds, giving time to press
the reset button on the Arduino.

\subsection{ESP32 Firmware Flashing Using ArduinoDroid}

The ESP32 firmware starts a TCP server on port~328 and relays bytes between
the WiFi socket and UART2.
Code: \texttt{esp32\_programmer/src/main.cpp}.
Replace \texttt{YOUR\_SSID} and \texttt{YOUR\_PASSWORD} with your network
credentials.

\begin{enumerate}
\item Compile using PlatformIO.
\item Connect the ESP32 to the phone via OTG cable.
\item Open ArduinoDroid $\rightarrow$ paste firmware $\rightarrow$ select
      \textit{ESP32 Dev Module}.
\item Tap Upload (hold BOOT on the ESP32 if upload fails).
\item Use Serial Monitor at 115200 baud to read the IP address.
\end{enumerate}

\subsection{Arduino Project Setup}

\subsubsection{Create Project}
Use Commands
\begin{lstlisting}[language=bash, caption={Initialize Arduino project}]
mkdir arduino_code && cd arduino_code
pio init --board uno
nvim src/main.cpp  # write your sketch
\end{lstlisting}

\subsubsection{PlatformIO Configuration}
Paste the contents of \texttt{arduino\_code/platformio.ini} into your
project's \texttt{platformio.ini}.
Replace \texttt{ESP\_IP} with the ESP32's IP address
(e.g.\ \texttt{192.168.43.100}).
The configuration uses \texttt{upload\_protocol = custom} with \texttt{avrdude}
connecting over \texttt{socket://ESP\_IP:328} at 115200 baud, targeting
\texttt{atmega328p} with the \texttt{stk500v1} programmer.

\subsection{Uploading Firmware Wirelessly}
Use Commands
\begin{lstlisting}[language=bash, caption={Build and upload}]
cd arduino_code
pio run --target upload
\end{lstlisting}

Press RST on the Arduino immediately after running the command.

\subsection{Finding the ESP32 IP Address}

The ESP32 IP address changes when switching networks.
Connect via OTG and open the ArduinoDroid Serial Monitor at 115200 baud,
then press RST on the ESP32 to read the assigned IP.

\subsection{The Timing Issue}

Uploading relies on the Arduino auto-reset mechanism.
When a serial connection opens, the DTR line drops, pulsing the RESET pin and
forcing the Arduino into bootloader mode.
The bootloader listens for only $\approx$1\,second before timing out.
Over TCP, \texttt{avrdude} cannot control DTR automatically.
The 100\,\textmu F capacitor (Section~II-C3) solves this by extending the
window to 1--3\,seconds.

%---------------------------------------------------------------------
\section{ESP32-to-ESP32 OTA Firmware Updates}
%---------------------------------------------------------------------

\subsection{Introduction}

This chapter extends the wireless programming concept to ESP32-to-ESP32
over-the-air (OTA) firmware updates.
Instead of a WiFi-to-UART bridge, ESP32~\#1 acts as a WiFi Access Point and
HTTP OTA controller.
ESP32~\#2 connects to it and updates its own firmware wirelessly---no USB
required after initial setup.

\textbf{Method:} HTTP OTA over raw TCP is used instead of Arduino OTA.
Arduino OTA relies on mDNS discovery and has no built-in integrity
verification, making it unreliable for ESP-to-ESP flashing.
HTTP OTA uses direct IP addressing with MD5 checksum verification.

\subsection{System Architecture}

\begin{table}[H]
\centering
\caption{Device Roles and Addresses}
\resizebox{\columnwidth}{!}{%
\begin{tabular}{llll}
\toprule
\textbf{Device} & \textbf{Role} & \textbf{IP} & \textbf{Port} \\
\midrule
ESP32 \#1 & Sender / Access Point & 192.168.4.1 & 80 \\
ESP32 \#2 & Receiver / OTA Target & 192.168.4.2 & 8080 \\
\bottomrule
\end{tabular}}
\end{table}

\subsubsection{Network Topology}
ESP32~\#1 operates as a WiFi access point with SSID
\texttt{ESP32-OTA-Net} (password \texttt{12345678}).
ESP32~\#2 connects to this network as a station.

\subsubsection{Data Flow}
\begin{enumerate}
\item Phone POSTs \texttt{firmware.bin} to ESP32~\#1 at \texttt{/upload}.
\item ESP32~\#1 saves binary to LittleFS and computes MD5.
\item Phone GETs \texttt{/trigger} on ESP32~\#1.
\item ESP32~\#1 opens raw TCP connection to \texttt{192.168.4.2:8080}.
\item ESP32~\#1 sends HTTP POST with binary body and \texttt{x-md5} header.
\item ESP32~\#2 streams bytes into the Update library, verifies MD5.
\item ESP32~\#2 replies HTTP 200 OK.
\item ESP32~\#2 calls \texttt{Update.end(true)} $\rightarrow$ sets boot
      partition $\rightarrow$ \texttt{ESP.restart()}.
\end{enumerate}

\subsection{Initial Setup}

\subsubsection{Flash ESP32 \#2 (Receiver) via USB}
Compile the following code using PlatformIO:
\texttt{esp32\_receiver/src/main.cpp}.
Using ArduinoDroid, flash via USB.
This is the only time USB is needed for ESP32~\#2.
On boot it connects to \texttt{ESP32-OTA-Net} and starts the OTA server on
port~8080.

\subsubsection{Flash ESP32 \#1 (Sender) via USB}
Compile the following using PlatformIO:
\texttt{esp32\_sender/src/main.cpp}.
Using ArduinoDroid, flash via USB.
ESP32~\#1 boots as WiFi AP with SSID \texttt{ESP32-OTA-Net},
password \texttt{12345678}, IP \texttt{192.168.4.1}.

\subsubsection{Verify Both Boards}
Connect your phone to \texttt{ESP32-OTA-Net}, then:

\begin{lstlisting}[language=bash, caption={Verify both boards}]
curl http://192.168.4.1/status      # ESP32 #1
curl http://192.168.4.2:8080/status # ESP32 #2
\end{lstlisting}

Both should respond with \texttt{\{"status":"ready",...\}}.

\subsection{Repeatable Update Workflow}

\subsubsection{Step 1 --- Compile Firmware Binary}
Use Commands
\begin{lstlisting}[language=bash, caption={Compile with PlatformIO}]
pio run
# binary at: .pio/build/esp2/firmware.bin
\end{lstlisting}

\subsubsection{Step 2 --- Upload Binary to ESP32 \#1}
Use Commands
\begin{lstlisting}[language=bash, caption={Upload binary to ESP32 \#1}]
# PlatformIO binary
curl -F "firmware=@.pio/build/esp2/firmware.bin" \
     http://192.168.4.1/upload
# Verify it landed
curl http://192.168.4.1/status  # firmware_loaded: true
\end{lstlisting}

\subsubsection{Step 3 --- Trigger OTA Push}
Use Commands
\begin{lstlisting}[language=bash, caption={Trigger OTA push}]
curl http://192.168.4.1/trigger
\end{lstlisting}

ESP32~\#1 connects to \texttt{192.168.4.2:8080}, sends the binary with the
MD5 header, and reports success.
ESP32~\#2 verifies, writes, and reboots.

\subsubsection{Step 4 --- Verify}
Wait $\approx$5\,seconds for ESP32~\#2 to reboot, then:
\begin{lstlisting}[language=bash, caption={Verify ESP32 \#2 is back online}]
curl http://192.168.4.2:8080/status
\end{lstlisting}

\subsection{Golden Rule}

Every sketch deployed to ESP32~\#2 \textbf{must} include
\texttt{handleOTAClient()} and \texttt{otaServer.available()} in
\texttt{loop()}.
If you flash a plain sketch without it, ESP32~\#2 loses wireless update
capability and requires USB to recover.
See the receiver template for the minimum required structure.

\begin{lstlisting}[language=C++, caption={Minimum required structure for every ESP32 \#2 firmware}]
#include <WiFi.h>
#include <Update.h>

WiFiServer otaServer(8080);
// paste handleOTAClient() from esp32_receiver_final.ino

void setup() {
  connectWiFi();     // connect to ESP32-OTA-Net
  otaServer.begin(); // always start OTA server
  // your app setup ...
}

void loop() {
  WiFiClient c = otaServer.available();
  if (c) handleOTAClient(c); // always poll for OTA
  // your app code below ...
}
\end{lstlisting}

%---------------------------------------------------------------------
\section{Conclusion}
%---------------------------------------------------------------------

ESP-Flasher demonstrates that a fully USB-free firmware deployment chain is
achievable from an Android phone running Termux.
Chapter~I establishes a reliable WiFi-to-UART bridge for AVR targets by
characterising and solving the DTR timing problem inherent in socket-based
\texttt{avrdude} operation.
Chapter~II extends the concept to ESP32 targets through a direct-IP HTTP OTA
scheme that requires no cloud infrastructure or mDNS, operating entirely over
a self-hosted access-point network.
The combined system provides a reproducible, command-line-only workflow
suitable for constrained development environments where USB host access is
unavailable.

%---------------------------------------------------------------------
\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
